Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-05-05 causal-bandit/nips_2016/general.tex
changes theorem 3
makes things clearer and more concise

For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)  
} 
Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.

We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).

@@

To address this difficult problem, we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$. $\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$. We then assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known. These could be estimated from experimental data on the same covariates $\boldsymbol{X}$, but where the outcome of interest differed such that $Y$ was not included, or similarly from observational data subject to identifiability constraints.

@@

\begin{figure}
    \begin{subfigure}[b]{0.3\textwidth}
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (4)
    (2) edge (4);
\end{tikzpicture}
        \caption{}
        \label{fig:causalStructure_confounded}
    \end{subfigure}
    \begin{subfigure}[b]{0.69\textwidth}
         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, right=of 4](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
  	(2) edge (3)
    (3) edge (4)
    (4) edge (5);
\end{tikzpicture}
        \caption{}
        \label{fig:causalchain}
    \end{subfigure}
    \caption{Example causal graphs}\label{fig:animals}
\end{figure}

Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.

\begin{wrapfigure}[14]{r}{0.5\textwidth}
\vspace{-12pt}
\begin{minipage}{.5\textwidth}
\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t = a$ with probability $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
}
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{wrapfigure}

@@

The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimized to minimize the worst-case simple regret. Abusing notation we define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
}
We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.

\begin{theorem}\label{thm:general}
If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by
\eq{
B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.
}
Then for $C = \sqrt{2} + 4$ we have
\eq{
\mu^* - \EE[\mu_I] \leq C\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}\,.
}
\end{theorem}

\subsection*{Choosing the Sampling Distribution}

Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$. In light of Theorem \ref{thm:general}
a natural choice of $\eta$ is the minimiser of $m(\eta)$.
\eq{
\eta^* 
= \argmin_\eta m(\eta) 
&= \argmin_\eta \max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
&= \argmin_\eta \underbrace{\max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
}
Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
Therefore the minimisation problem may be tackled using standard techniques from convex optimisation. 
An interpretation of $m(\eta^*)$ is the minimum achievable worst-case variance of the importance weighted estimator.
In the experimental section we present some special cases,
but for now we give two simple results. The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.

@@

\begin{remark}
The algorithm in the previous section for the parallel bandit problem did not assume knowledge of the conditional distributions on $X$ and 
instead only the graph structure. Thus the algorithm given in that setting has broader applicability to that particular graph than the generic algorithm given here.
Finally, we believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
for future work.
\end{remark}

>>>>

For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
} 
where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.

We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, but not from $do()$.

@@

To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, or similarly from observational data subject to identifiability constraints. Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.

%we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$. $\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$. We then

@@

\begin{wrapfigure}[14]{r}{0.5\textwidth}
\vspace{-12pt}
\begin{minipage}{.5\textwidth}
\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t = a$ with probability $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\FOR{$a \in \calA$}
\STATE
\eq {
\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
}
\ENDFOR
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{wrapfigure}

@@

The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimized to minimize the worst-case simple regret. Abusing notation we define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
}

We will show shortly that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.

\begin{theorem}\label{thm:general}
If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by $B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.$

\eq{
\simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}}\,.
}
\end{theorem}

The proof is in the supplementary materials. Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$. Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$. Thus it has broader applicability to the parallel graph than the generic algorithm given here. We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
for future work.

\paragraph{Choosing the Sampling Distribution} Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$. In light of Theorem \ref{thm:general}
a natural choice of $\eta$ is the minimiser of $m(\eta)$.
\eq{
\eta^* 
= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
}
Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
Therefore the minimisation problem may be tackled using standard techniques from convex optimisation. An interpretation of $m(\eta^*)$ is the minimum achievable worst-case variance of the importance weighted estimator. In the experimental section we present some special cases, but for now we give two simple results. The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.

@@

### 2016-05-05 causal-bandit/nips_2016/introduction.tex
makes things clearer and more concise

The type of problem we are concerned with is best illustrated with an example. 
Consider a farmer wishing to optimise the yield of her crop. 
She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
In each season the farmer has enough time and money to intervene and control at most one of these variables:
deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
%without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?

\subsection{Contributions}

We take the first step towards formalising and solving problems such as the one above. 
In \S\ref{sec:defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph. 
We show that our causal bandit framework subsumes the classical bandits (no additional observations) and contextual stochastic bandit problems (observations are revealed before an intervention is chosen) before focusing on the case where, like the above example, observations occur \emph{after} each intervention is made.

@@

\subsection{Related Work}

As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.

>>>>

The type of problem we are concerned with is best illustrated with an example. 
Consider a farmer wishing to optimise the yield of her crop. 
She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
In each season the farmer has enough time and money to intervene and control at most one of these variables:
deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
%without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?
\paragraph{Contributions} We take the first step towards formalising and solving problems such as the one above. 
In \S\ref{sec:defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph. 
We show that our causal bandit framework subsumes the classical bandits (no additional observations) and contextual stochastic bandit problems (observations are revealed before an intervention is chosen) before focusing on the case where, like the above example, observations occur \emph{after} each intervention is made.

@@

\paragraph{Related Work} As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round. In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}. However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.

### 2016-05-05 causal-bandit/nips_2016/setup.tex
makes things clearer and more concise

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
We will assume each variable only takes on a finite number of distinct values.
An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from $\parents{i}$ to $X_i$ for each $X_i \in \vec{X}$. 
The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Details can be found in Chapter 21 of~\cite{Koller2009}.

@@

The objective of the learner is to minimise the simple regret
\begin{equation}
\label{eq:regret-simple}
	\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.
\end{equation}
This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and
is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefintely.

@@

We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one 
edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some 
arbitrary but unknown, real-valued function $r$. The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by 
treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
Intuitively though, one would expect to perform better by making use of the extra observations.

>>>>

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
We will assume each variable only takes on a finite number of distinct values.
An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from $\parents{i}$ to $X_i$ for each $X_i \in \vec{X}$. 
The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Details can be found in Chapter 21 of~\cite{Koller2009}.

@@

The objective of the learner is to minimise the simple regret $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$ This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefinitely.

@@

We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$. The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$. Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward. Intuitively though, one would expect to perform better by making use of the extra structure and observations.
